---
title: "Automated EDA"
author: "YPhillips"
date: "2024-03-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###  title: "Automated Exploratory Data Analysis"

```{r}
library(rmarkdown)
library(magrittr) #uses the pipe operator, %>% 
library(dplyr) #glimpse function

options(scipen=999,digits=3) #scientific notation
```

# Set working directory. Working directory is a default location on your machine where items are saved.
#Point and click: Sessions -> Choose Working Directory -> Set Working Directory. Then naviagte to the folder on your machine were you want items saved.

# Place file location here: 
#example: setwd("C:/Users/redbe/OneDrive/Documents/Data")

# import .csv or .txt data files 
# USe either 'Import Dataset' button (top right), Use 'From Text (base)' or the script to import a .csv file. 

```{r}
purchase_order <- read.csv("purchase-order-data-2012-2015.csv")
```


```{r}
View(purchase_order) #View data in the Source window
```

# Confirm the dimensions. Dimension are the number of (rows,columns).
```{r}
dim(purchase_order)
```

# Confirm data types of the variables
```{r}
str(purchase_order)
```

#removing columns
#use the matrix[row,column] syntax
```{r}
purchase_order <- purchase_order[,-8] #8 is the position of the of variable in the dataset. For multiple columns, use the setup purchase_order[,-c(8,15)] to remove #8 and #15.
```

#Data Types ---------
#There are numerous data types in R that store various kinds of data.

# Numeric data

#numeric data is the most common type in R. The numeric data type handles integers and decimals, both positive and negative, and zero. A numeric value stored in a variable is automatically assumed to be numeric.

#Another important numeric is the type integer. The integer is for whole numbers only, no decimals. To set an integer to a variable it is necessary to append the value with an L. 

# Character data

#The character (string) data type is very common in statistical analysis. R has two primary ways of handling character data: character and factor

#Characters are case sensitive, so “Data” is different from “data” or “DATA”. To find the length of a character (or numeric) use the nchar function. But this will not work for factor data.

# Dates
#Dealing with dates and times can be difficult. The most useful are Date and POSIXct. Date stores just a date while POSIXct stores a date and time. Both objects are actually represented as the number of days (Date) or seconds (POSIXct) since January 1, 1970.


# As.(dot) functions - changes from one data type to another data type

# To change to character, use the following:
#dataset$columnname <- as.character(dataset$columnname)
```{r}
purchase_order$Supplier.Code <- as.character(purchase_order$Supplier.Code)
purchase_order$Classification.Codes <- as.character(purchase_order$Classification.Codes)
purchase_order$Normalized.UNSPSC <- as.character(purchase_order$Normalized.UNSPSC)
```

# preprocess data: example includes removing the dollar sign '$' from the dataset in Unit.Price and Total.Price. You can do this in Excel or use script:

# Creating a function - stored in the Environment window
# To change character to numeric, use the following:
#dataset$columnname <- as.numeric(dataset$columnname)

```{r}
# Function to clean and convert character columns to numeric
convert_to_numeric <- function(data, columns) {
  for (col in columns) {
    data[[col]] <- gsub("\\$", "", data[[col]])   # Remove dollar signs
    data[[col]] <- gsub(",", "", data[[col]])    # Remove commas
    data[[col]] <- gsub("\\s+", "", data[[col]]) # Remove spaces
    data[[col]] <- as.numeric(data[[col]])       # Convert to numeric
  }
  return(data)
}

# Apply the function
columns_to_convert <- c("Unit.Price", "Total.Price")
purchase_order <- convert_to_numeric(purchase.order.data.2012.2015, columns_to_convert)

# Output the cleaned data
print(purchase_order)
str(purchase_order)
```

# To change numeric to integer, use the following:
#dataset$columnname <- as.integer(dataset$columnname)

# To change character to factor (nominal). A factor is a categorical variable. use the following:
```{r}
purchase_order$Acquisition.Type <- factor(purchase_order$Acquisition.Type)
purchase_order$Sub.Acquisition.Type <- factor(purchase_order$Sub.Acquisition.Type)
purchase_order$Acquisition.Method <- factor(purchase_order$Acquisition.Method)
purchase_order$Family <- factor(purchase_order$Family)
purchase_order$Segment.Title <- factor(purchase_order$Segment.Title)
purchase_order$Fiscal.Year <- factor(purchase_order$Fiscal.Year)
purchase_order$Class <- factor(purchase_order$Class)
purchase_order$Segment <- factor(purchase_order$Segment)
```

# To change character to factor (ordinal), use the following:
#dataset$columnname <- factor(dataset$columnname, levels,ordered = TRUE)

# To change character to data, use the following:
# Convert Purchase.Date to Date format
```{r}
purchase_order$Purchase.Date <- as.Date(purchase_order$Purchase.Date, format = "%m/%d/%Y")
```

#reconfirm data types are how you want them.
```{r}
str(purchase_order)
```

#First thing to be noticed is that there is 100% complete rate. From observation of purchase_order dataset this can not be possible. 

# If you have missing variables not coded NA. Change the character (i.e ?) to "NA" missing values in an entire dataset (dataframe)
```{r}
purchase_order[purchase_order == "-"] <- NA #ex. the dash,-, used for missing
```

# Determine which columns return column names containing missing observations. Only change the dataframe name in the script below.
```{r}
list_na <- colnames(purchase_order)[apply(purchase_order, 2, anyNA) ]
list_na
```

```{r}
summary(purchase_order$Unit.Price)
```

# Look into dropping variables instead of removing obs until more is known about the data set
# which variable(s) to drop; missing variables with >50% of obs

# Automatic EDA

# The dataMaid package creates a report in different formats, such as PDF, DOCX, or HTML. The generated report checks and neatly simply summarizes the dataset. It is a good tool for checking errors in the dataset. Additional information https://cran.r-project.org/web/packages/dataMaid/index.html

# Install package
```{r}
#install.packages("dataMaid")
```

# Import (load) library in current session
```{r}
library(dataMaid)
# Create report
makeDataReport(purchase_order, output = "html", replace = TRUE)
```

# DataExplorer, we can simplify the EDA process and report generation. This library automatically scans the dataset for variables, performs data profiling, and provides many useful functions to create various charts on both discrete and continuous features in the dataset. The DataExplorer library generates a complete HTML report in the working directory for the EDA

# Install package
```{r}
#install.packages("DataExplorer")
```

# Import (load) library in current session
```{r}
library(DataExplorer)
create_report(purchase_order, 
              report_title = "Custom EDA Report",
              output_file = "
              ")
```

#The SmartEDA library also allows us to generate a complete HTML report using the ExpReport function. Let us install and import the package as well as run the ExpReport function to perform the EDA, as shown below.

# Install package
```{r}
#install.packages("SmartEDA")
```

# Import library
```{r}
library(SmartEDA)
ExpReport(purchase_order, op_file='smartEDA.html')
```

# Install package
#install.packages("arsenal")

# Install package
install.packages("arsenal")

# quantiative analysis
# descriptive stats 
```{r}
summary(purchase_order)
```

# qualitative analysis
# cross tabulations 

# Install package
```{r}
#install.packages("flextable")
#install.packages("crosstable")
```

```{r}
library(flextable)
library(crosstable)

crosstable(purchase_order, c(Total.Price, Quantity), by = Acquisition.Type) %>% #then   
  as_flextable(keep_id=TRUE)
```

# Frequency table of categorical variable, sorted decreasing order
```{r}
View(sort(table(purchase_order$Segment.Title), decreasing =TRUE))
```

# Additional Resources
https://bookdown.org/ageraci/STAT160Companion/
https://rc2e.com/
